Evolutionary stretegies

### 2016-01-23 evolutionaryComputation/summery.tex
finished page 1 of the topics (excluding self knowledge)

% This sets page margins to .5 inch if using letter paper, and to 1cm
% if using A4 paper. (This probably isn't strictly necessary.)
% If using another size paper, use default 1cm margins.
\ifthenelse{\lengthtest { \paperwidth = 11in}}
    { \geometry{top=.5in,left=.5in,right=.5in,bottom=.5in} }
    {\ifthenelse{ \lengthtest{ \paperwidth = 297mm}}
        {\geometry{top=1cm,left=1cm,right=1cm,bottom=1cm} }
        {\geometry{top=1cm,left=1cm,right=1cm,bottom=1cm} }
    }

@@

\section{Genetic algorithms}
%TODO: crossover methods I don't understand
%TODO: schemas, buidling block theory
\subsection{Schema}
$o(h)$: schema order $o(11\#\#0)$ ie nonwildcard \# elements.
$\delta(h)$: schema defining length, maximum distance between two defining
symbols. $m(h,t)$: number of schema h instances at generation t.
$f(h,t)=\sum_{i\in P}f_i$ schema fitness is average fitness of members.
\subsubsection{Growth by selection}
reproduction ratio $\phi(h,t)=\frac{m(h,t^s)}{m(h,t)}$.
\paragraph{Proportionate selection}
$f_i$ fitness ind i. Probability individiual i selected: $\frac{f_i}{\sum f_i}$.
$N$ pop size. Expected number of copies
$\frac{f_i}{\sum f_i} \cdot N = \frac{f_i}{f(t)}$.
Expected number of copies of schema $h$ members:
$m(h,t^s) = m(h,t)\phi(h,t)=m(h,t)\frac{f(h,t)}{f(t)}$.
\paragraph{Tournament selction}
tournament size $s: 0 \leq \phi(h,t) \leq s$.

@@

\subsubsection{Theorom}
Selection mutation and recombination combined:
$m(h,t+1)\leq m(h,t)\phi(h,t)[1-\epsilon(h,t)]$.
Net growth factor: $\gamma(h,t)=\frac{m(h,t+1)}{m(h,t)}$.

\subsection{Fitness Correlation coefficents}
$F_p$: mean fitness of the parents. $F_c$: mean fitness of the children.
$\sigma(F_p) =$ standard deviation of fitness parents
$\sigma(F_c) =$ std. dev. of children. coveriance fitness between parents and children:
$cov(F_p,F_c)=\sum^n_{i=1}\frac{(f(p_{gi})-F_p)(f(c_{gi})-f_c)}{n}$.
Operator fitness correlation coeficient
$\rho_{op}=\frac{cov(F_p,F_c)}{sigma(F_p)\sigma(F_c)}$.

\section{ES representation}
Fitness function: $f(x_1,\dots,x_n):\mathbb{R}^n\to \mathbb{R}$.
Genotype representation of an individual solution:
$(x_1,\dots,x_n,\sigma^2_1,\dots,\sigma^2_n, c_{12},\dots,c_{n-1,n})$,
paremeters $(x_1, \dots,x_n)$ need to be optimized.
Individiual solution consists of 3 parts: $\vec{x}$: problem vairables
$\Rightarrow$ Fitness $f(\vec{x})$. $\vec{sigma}$: standard deviations
$\Rightarrow$ variances. $\vec{\alpha}$: rotation angles $\Rightarrow$
covariances. The $n$-dimensional normal probability density function:
$p(X=x_1,\dots,x_n)=\frac{exp(-\frac{1}{2}X^TC^{-1}X)}{\sqrt{(2\pi)^n|C|}}$.
$C$: correlation matrix ($c_{ij}$; $|C|$ determinant $\Rightarrow$ rotation 
angles $\alpha_{ij}$: tan $2\alpha_{ij}=2c_{ij}/(\sigma^2_i-\sigma^2_j)$.
1-dimensional gaussian function:
$f(x)=\frac{1}{\sqrt{2\pi \sigma^2}}exp\frac{(x-\mu)^2}{2\sigma^2}$
\subsection{Mutation}
Complete covairance matrix: $\sigma_1,\dots,\sigma_n$; $c_{ij}\neq 0 (i\neq j)$.
The strategy paremeters are mutated:
$\sigma'_i=\sigma_ie^{
	\frac{\mathbb{N}(0,1)}{\sqrt{2n}}+
	\frac{\mathbb{N}_i(0,1)}{\sqrt{2\sqrt{n}}}
}$
$\alpha'_j=\alpha_j+\beta\mathbb{N}_j(0,1)$.
$\beta = 5$ degree in radians. $\mathbb{N}(0,1)$: standard normal
distribution. Then the problem paremeters are mutated with the new
strategy: $\vec{x}'=\vec{x}+\vec{N}(\vec{0},\vec{\sigma},\vec{\alpha}')$.
$\vec{N}$: n-dimensional normal distribution.
Often this is simplified by not doing angles or even just having a single
standard deviation (ie no index at $\sigma$ and nothing after the + 
in the power $e$).
\subsection{Recombination}
create one offspring from several parents selected randomly. Problem parameters
are averaged. The std devation is selected from one of the parents. Rotation
angles are not recombined
\subsection{Selection}
ES applies a high selection pressure: from $\mu$ parents $\lambda$ offspring
are generated with $\lambda >> \mu$ (~5 to 10 times). The best $\mu$ solutions
of the $\lambda$ offspring are selected, (or from both offspring and parents,
but experiments show just offspring has better selfadaptation).
\section{Ada pursuit}
Given: Set of $K$ \emph{operators} $A=\{a_1,\dots,a_k\}$. 
\emph{Probability vector} $P(t) = \{P_1(t),\dots,P_k(t)\}$. operator $a_i$ 
applied at time t in proportion to probability $P_i(t)$. env. returns rewards
$R_i(t) \geq 0$. Goal: Adapt $P(t)$ so that expected value of cumalitive
reward $\varepsilon[R]=\sum^T_{t=1}R_i(t)$ is maximized.

>>>>

% This sets page margins to .5 inch if using letter paper, and to 1cm
% if using A4 paper. (This probably isn't strictly necessary.)
% If using another size paper, use default 1cm margins.
\ifthenelse{\lengthtest { \paperwidth = 11in}}
    { \geometry{top=.5in,left=.5in,right=.5in,bottom=.5in} }
    {\ifthenelse{ \lengthtest{ \paperwidth = 297mm}}
        {\geometry{top=0.5cm,left=0.5cm,right=0.5cm,bottom=0.5cm} }
        {\geometry{top=0.5cm,left=0.5cm,right=0.5cm,bottom=0.5cm} }
    }

@@

\section{Genetic algorithms}
%TODO: crossover methods I don't understand
%TODO: schemas, buidling block theory
\subsection{Schema}
$o(h)$: schema order $o(11\#\#0)$ ie nonwildcard \# elements.
$\delta(h)$: schema defining length, maximum distance between two defining
symbols. $m(h,t)$: number of schema h instances at generation t.
$f(h,t)=\sum_{i\in P}f_i$ schema fitness is average fitness of members.

\emph{Buidling blocks} low order high performance schemate receive exponentiaonally
(geometrically) increasing trials.

\subsubsection{Growth by selection}
reproduction ratio $\phi(h,t)=\frac{m(h,t^s)}{m(h,t)}$.

\emph{Proportionate selection}
$f_i$ fitness ind i. Probability individiual i selected: $\frac{f_i}{\sum f_i}$.
$N$ pop size. Expected number of copies
$\frac{f_i}{\sum f_i} \cdot N = \frac{f_i}{f(t)}$.
Expected number of copies of schema $h$ members:
$m(h,t^s) = m(h,t)\phi(h,t)=m(h,t)\frac{f(h,t)}{f(t)}$.

\emph{Tournament selction}
tournament size $s: 0 \leq \phi(h,t) \leq s$.

@@

\subsubsection{Theorem}
Selection mutation and recombination combined:
$m(h,t+1)\leq m(h,t)\phi(h,t)[1-\epsilon(h,t)]$.
Net growth factor: $\gamma(h,t)=\frac{m(h,t+1)}{m(h,t)}$.
\subsection{Permutation problems}
2-opt m, select 2 points and reverse a subsequance.
adjacency of elements is important, 2-opt has minimal change.
relative ordering is important: 2-opt large change.
\subsubsection{Crossover}
std. crosover won't work cause they can make invalid solution.
ie : $abcd+CBDA\to Cbcd$. Order crossover: rnd select 2 points,
copy subseq between crosspoints from p1, starting at 2nd crosspoint
fill in missing elements retaining order from p2. partially
mapped crosover: rnd select 2 points, copy p2 to child, copy
elements between crospoint from p1 to childe while placing
the replaced element from p2 at the location where the replacer is.
Position crossover: rnd select k postitions, copy unmarked elements from
p1 to child, scan p2 from lft to rght and fill in missing elemnts. Max 
preservative cross: rnd select two points, copy subseq between crospoints
from p1, add succesivily an adjecent element from p2 starting at
last element in child, if already placed take and adjecent elemnt form
p1. Cycle crossover: mark cycles, cross full cycles.

\subsection{Fitness Correlation coefficents}
Genetic operators should perserve usefull fitness characaristics.
Calcualte the fitness coorlation coeficient to quantify this.
$F_p$: mean fitness of the parents. $F_c$: mean fitness of the children.
$\sigma(F_p) =$ standard deviation of fitness parents
$\sigma(F_c) =$ std. dev. of children. coveriance fitness between parents and children:
$cov(F_p,F_c)=\sum^n_{i=1}\frac{(f(p_{gi})-F_p)(f(c_{gi})-f_c)}{n}$.
Operator fitness correlation coeficient
$\rho_{op}=\frac{cov(F_p,F_c)}{sigma(F_p)\sigma(F_c)}$.

\section{Metaheursitics}
Crossover can have mutliple search biases:
1. Random sampling witin a specific subspace. 2. place together for contrast
partial solutions from two parent solutions.

\subsection{PBMS}
Probablistic Modal Building: Probability distributions model dependencies 
between variables present in good solutions. sSelection makes these fitness 
based dependencies stand out. Estimating a probablity modal over the selected
solutions identifies these dependencies. Drawing new samples form the probability
modal will respect the dependencies.
PMBLS = PBM + LS. Bivariate probalbistic modal learns the pairwise dependencies
between problem dependencies. Buildign a dependency tree=maximum spanning tree
over the dependency graph. New solutions obtained by sampling from the dependency
tree.
For graphbipartitioning. Probablistic model: count the frequnceis that two
verticies are in the same parition. Dep tree build over most extreme frequency
values. reduce computational complexity by only considering the
pairwise interactions between connected vertices
\section{ES representation}
Fitness function: $f(x_1,\dots,x_n):\mathbb{R}^n\to \mathbb{R}$.
Genotype representation of an individual solution:
$(x_1,\dots,x_n,\sigma^2_1,\dots,\sigma^2_n, c_{12},\dots,c_{n-1,n})$,
paremeters $(x_1, \dots,x_n)$ need to be optimized.
Individiual solution consists of 3 parts: $\vec{x}$: problem vairables
$\Rightarrow$ Fitness $f(\vec{x})$. $\vec{sigma}$: standard deviations
$\Rightarrow$ variances. $\vec{\alpha}$: rotation angles $\Rightarrow$
covariances. The $n$-dimensional normal probability density function:
$p(X=x_1,\dots,x_n)=\frac{exp(-\frac{1}{2}X^TC^{-1}X)}{\sqrt{(2\pi)^n|C|}}$.
$C$: correlation matrix ($c_{ij}$; $|C|$ determinant $\Rightarrow$ rotation 
angles $\alpha_{ij}$: tan $2\alpha_{ij}=2c_{ij}/(\sigma^2_i-\sigma^2_j)$.
1-dimensional gaussian function:
$f(x)=\frac{1}{\sqrt{2\pi \sigma^2}}exp\frac{(x-\mu)^2}{2\sigma^2}$
\subsection{Mutation}
Complete covairance matrix: $\sigma_1,\dots,\sigma_n$; $c_{ij}\neq 0 (i\neq j)$.
The strategy paremeters are mutated:
$\sigma'_i=\sigma_ie^{
	\frac{\mathbb{N}(0,1)}{\sqrt{2n}}+
	\frac{\mathbb{N}_i(0,1)}{\sqrt{2\sqrt{n}}}
}$
$\alpha'_j=\alpha_j+\beta\mathbb{N}_j(0,1)$.
$\beta = 5$ degree in radians. $\mathbb{N}(0,1)$: standard normal
distribution. Then the problem paremeters are mutated with the new
strategy: $\vec{x}'=\vec{x}+\vec{N}(\vec{0},\vec{\sigma},\vec{\alpha}')$.
$\vec{N}$: n-dimensional normal distribution.
Often this is simplified by not doing angles or even just having a single
standard deviation (ie no index at $\sigma$ and nothing after the + 
in the power $e$).
\subsection{Recombination}
create one offspring from several parents selected randomly. Problem parameters
are averaged. The std devation is selected from one of the parents. Rotation
angles are not recombined
\subsection{Selection}
ES applies a high selection pressure: from $\mu$ parents $\lambda$ offspring
are generated with $\lambda >> \mu$ (~5 to 10 times). The best $\mu$ solutions
of the $\lambda$ offspring are selected, (or from both offspring and parents,
but experiments show just offspring has better selfadaptation).
\section{Ada pursuit}
Given: Set of $K$ \emph{operators} $A=\{a_1,\dots,a_k\}$. 
\emph{Probability vector} $P(t) = \{P_1(t),\dots,P_k(t)\}$. operator $a_i$ 
applied at time t in proportion to probability $P_i(t)$. env. returns rewards
$R_i(t) \geq 0$. Goal: Adapt $P(t)$ so that expected value of cumalitive
reward $\varepsilon[R]=\sum^T_{t=1}R_i(t)$ is maximized.
